---
title: "Storm Report"
author: "Michael Bristow"
date: "29 November 2017"
output: html_document
---

#Reproducible Research - Project 2
##1. Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

In this report,effect of weather events on personal as well as property damages was studied. Barplots were plotted seperately for the top 8 weather events that causes highest fatalities and highest injuries. Results indicate that most Fatalities and injuries were caused by Tornados.Also, barplots were plotted for the top 8 weather events that causes the highest property damage and crop damage.

##2. Data Processing
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size.

```{r cache=TRUE}
library(ggplot2)
#First, check if data downloaded, download and unzip
if(!file.exists('repdata_data_StormData.csv')){
    unzip('repdata_data_StormData.csv.bz2')
}
stormdata <- read.csv("./repdata_data_StormData.csv.bz2")

#Only uise the columns we need
reduceddata <- stormdata[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
```

##2.1 Data Subsetting and Aggregating
##2.1.1
Select only the top 10 Injuries and Deaths  
Select only the top 10 Property and Crop damage
```{r}
sumInjuries <- aggregate(INJURIES ~ EVTYPE, data = reduceddata, sum, na.rm = TRUE)
sumDeaths <- aggregate(FATALITIES  ~ EVTYPE, data = reduceddata, sum, na.rm = TRUE)

#get top 10
topInjuries <- sumInjuries[order(-sumInjuries$INJURIES), ][1:10, ]
topDeaths <- sumDeaths[order(-sumDeaths$FATALITIES), ][1:10, ]
```

```{r}
#sort out the cost of property damage fields
#replace character with correct 000's
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "K" | reduceddata$PROPDMGEXP == "3"] <- 1000
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "M" | reduceddata$PROPDMGEXP == "m" | reduceddata$PROPDMGEXP == "6"] <- 1e+06
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "" | reduceddata$PROPDMGEXP == "0"] <- 1
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "B"] <- 1e+09
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "5"] <- 1e+05
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "4"] <- 10000
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "2" | reduceddata$PROPDMGEXP == "h" | reduceddata$PROPDMGEXP == "H"] <- 100
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "7"] <- 1e+07
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "1"] <- 10
reduceddata$PROPMultiplier[reduceddata$PROPDMGEXP == "8"] <- 1e+08

#replace  with correct 000's
reduceddata$PROPDMGCorrected <- reduceddata$PROPDMG * reduceddata$PROPMultiplier
#sort and get top 10
sumPropDamage <- aggregate(PROPDMGCorrected ~ EVTYPE, data = reduceddata, sum, na.rm = TRUE)
topDamage <- sumPropDamage[order(-sumPropDamage$PROPDMGCorrected), ][1:10, ]


#do same for crop damage
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "K" | reduceddata$CROPDMGEXP == "3"] <- 1000
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "M" | reduceddata$CROPDMGEXP == "m" | reduceddata$CROPDMGEXP == "6"] <- 1e+06
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "" | reduceddata$CROPDMGEXP == "0"] <- 1
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "B"] <- 1e+09
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "5"] <- 1e+05
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "4"] <- 10000
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "2" | reduceddata$CROPDMGEXP == "h" | reduceddata$CROPDMGEXP == "H"] <- 100
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "7"] <- 1e+07
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "1"] <- 10
reduceddata$CROPMultiplier[reduceddata$CROPDMGEXP == "8"] <- 1e+08

#replace  with correct 000's
reduceddata$CROPDMGCorrected <- reduceddata$CROPDMG * reduceddata$CROPMultiplier
#sort and get top 10
sumCropDamage <- aggregate(CROPDMGCorrected ~ EVTYPE, data = reduceddata, sum, na.rm = TRUE)
topCropDamage <- sumCropDamage[order(-sumCropDamage$CROPDMGCorrected), ][1:10, ]


```

##3. Results
##3.1 Top 10 Causes
The top injury cause is `r topInjuries$EVTYPE[1]` with `r format(topInjuries$INJURIES[1], scientific = FALSE)` incidents  
The top cause of fatality is `r topDeaths$EVTYPE[1]` with `r format(topDeaths$FATALITIES[1], scientific = FALSE)` fatalities  

```{r}
#plot the injuries and deaths
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0))
barplot(topInjuries$INJURIES, las = 3, names.arg = topInjuries$EVTYPE, main = "Injuries - top 10 causes", 
        ylab = "# Injuries", col = "blue")
barplot(topDeaths$FATALITIES, las = 3, names.arg = topDeaths$EVTYPE, main = "Fatalities - top 10 causes", 
        ylab = "# fatalities", col = "red")
```
  
##3.2 Top 10 cause of Property & Crop Damage

```{r}
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0))
barplot(topDamage$PROPDMGCorrected/(1000000000), las = 3, names.arg = topDamage$EVTYPE, main = "Property Damage - top 10 causes", 
        ylab = "Cost (Billions)", col = "blue")
barplot(topCropDamage$CROPDMGCorrected/(1000000000), las = 3, names.arg = topCropDamage$EVTYPE, main = "Crop Damage - top 10 causes", 
        ylab = "Cost (Billions)", col = "red")
```

##Conclusions
Tornados are the biggest cause of both Fatalities and Injuries
Floods caused the most damage to property while drought was the largets cause of crop damage